---
title: "ESM206_Lab9"
author: "Kirsten White"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(here)
library(broom) 

```

## Read in the data
```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```
## Get counts of lizards by species
```{r}
# Find counts by species 

lizards %>%
  count(spp) %>%
  arrange(-n) # arrange in decreasing order by the variable n that contains the counts; shows counts of each lizard species and how many times they appear in the data sets; we see that UTST is the most common (will continue on with this)
```
## Make a subset 
- Contain only: USTS lizards, observations with 3 highest observed populations (IBPE - grassland, RABB - rabbit mesquite, and EAST - east tarbush site), and where tail condition is recorded (w = whole or b = broken)

```{r}
utst <- lizards %>%
  filter(spp == "UTST") %>%
  filter(site %in% c("IBPE", "RABB", "EAST")) %>% # look for any match within this vector; check with unique(utst$site) in console
  filter(tail %in% c("B", "W")) # excludes missing values or n/a values
```

## Now, let's find counts and proportions of lizards with broken or whole tails at each site 
```{r}
utst %>%
  count(site, tail)
```

```{r}
tail_counts <- utst %>%
  janitor::tabyl(site, tail) # returns the same info but instead of long format, it is in a contingency table with counts at each intersection (good for chi squared) 

tail_proportions <- tail_counts %>%
  adorn_percentages() %>% # this is from the janitor package (janitor::adorn_percentages()); shows percentage of lizards at each site 
  adorn_pct_formatting((digits = 2)) %>% # rounds percentages to 2 digits
  adorn_ns() # includes actual counts along with the percentages; helpful to include when reporting contingency table info 
```
column to row names so that all we are left with is the contingency tabels with counts of tails for each site

## convert to contingency table 
```{r}
# use `column_to_rowname()` to do this

lizard_ct <- tail_counts %>%
  column_to_rownames(var = "site") # remove this as a variable; remove the column and have site names listed as row name info (not as their own variable anymore)
```


### Chi-square
- Asking, are tail breakage and site truly independent (site has no impact on tail breakage; if the case, we would expect to see similar counts/proportions of each tail break type between sites)
- running the test is easy!
```{r}
lizard_x2 <- chisq.test(lizard_ct) # give it entire contingency table 
lizard_x2

lizard_tidy <- broom::tidy(lizard_x2) # writes out the info in tidy format and allows us to refer to the values in in-line referencing
```
- is that probability (p val = 0.001456), yes we have sufficient evidence to reject the null and retain the alternative hyp that site and tail breakage are NOT independent; there is a significant association between site and tail breakage; site and lizard tail breakage are not independent
- There is a significant association (non-independence) between site and lizard tail breakage ($chi$^2^(`r lizard_tidy$parameter`) = `r lizard_tidy$statistic`, *p* = `r lizard_tidy$p.value`)
